---
title: "Newark_Cycle_Depth_Rank"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
### Step 1: Data Collection
# Read Lake Cycle excel sheet
library(readxl)
lake_xls_f <- './LakeCycleHomework_AnalysisExcel.XLS'
excel_sheets(lake_xls_f)

lake_xls <- read_excel(lake_xls_f, sheet='Input Data')

# Complete Interval I of Newark Basin
# Interval I = 2768-3370 m; 706 pts
nb_d <- data.frame(lake_xls[8:713,c(16,17,18)])
colnames(nb_d) <- c('Point', 'Depth (m)', 'Depth Rank')

# Depth Vec
dv <- as.numeric(nb_d$`Depth (m)`)
#diff(dv)
max_d <- max(dv)
min_d <- min(dv)
# window of depth
window <- max(dv)-min(dv)

dr <- as.numeric(nb_d$`Depth Rank`)
max_dr <- max(dr)
min_dr <- min(dr)

plot(nb_d$`Depth (m)`, nb_d$`Depth Rank`, xlab='Depth(meters)', 
     ylab='Depth Rank', type='l', col='darkblue', lwd=2,
     main='Input Data for Spectral Analysis'
)

```

# Step 2. De-mean the data
```{r}
#________________________________________________________________________________#
nb_d.avg <- mean(dr)
nb_d.dm <- as.numeric(nb_d$`Depth Rank`) - nb_d.avg

# plot
plot(nb_d$`Depth (m)`, nb_d$`Depth Rank`, xlab='Depth(meters)', 
     ylab='', type='l', col='darkblue', lwd=2,
     main='Demeaned Input Data(Red Curve)', ylim=c(-4,4)
)
abline(h=nb_d.avg, lty=2, col='darkblue')
lines(nb_d$`Depth (m)`, nb_d.dm, col='red')
abline(h=0, lty=2, col='red')
```

# Step 3. Filter the data

```{r}
# Filtering - White noise removed
# (3pt moving "Hanning" average)
# 0.5*n + 0.25*(n-1 + n+1)
hanning_ma <- function(df, n) {
  x <- 0.5 * df[n] + 0.25*(df[n-1] + df[n+1])
  return(x)
}

nb_d.filt <- c()
i <- 1
nr <- length(nb_d.dm)
for (n in 2:(nr-1)) {
  h <- hanning_ma(nb_d.dm, n)
  nb_d.filt <- c(nb_d.filt, h)
}

# plot
plot(nb_d$`Depth (m)`, nb_d.dm, xlab='Depth(meters)', 
     ylab='Value', type='l', col='red', lwd=1,
     main='Demeaned + Filtered Input Data', ylim=c(-4,4)
)
abline(h=0, lty=2, col='red')
lines((nb_d$`Depth (m)`)[2:(nr-1)], nb_d.filt, col='green')
```

# Step 4. Split-Cosine Tapering
```{r}
nb_d.tap <- nb_d.filt
i <- 1
nr <- length(nb_d$`Depth (m)`)
n <- length(nb_d.filt)

tap_perc_n <- n * 10/100 # 10% tapering to reduce the effect of the terminations of data at the ends of the window

for (j in 1:tap_perc_n) {
  w = j * pi / tap_perc_n
  nb_d.tap[j] = nb_d.filt[j] * 0.5 * (1-cos(w))
}
for (j in (n-tap_perc_n):n) {
  w = (512-j) * pi / tap_perc_n
  nb_d.tap[j] = nb_d.filt[j] * 0.5 * (1-cos(w))
}

# plot
plot(nb_d$`Depth (m)`, nb_d.dm, xlab='Depth(meters)', 
     ylab='', type='l', col='red', lwd=1,
     main='Demeaned + Filtered + Split cosine tapered(10%) Input Data', ylim=c(-8,8)
)
abline(h=0, lty=2, col='red')
lines((nb_d$`Depth (m)`)[2:(nr-1)], nb_d.filt, col='green')
lines((nb_d$`Depth (m)`)[2:(nr-1)], nb_d.tap, col='black')

# plot(nb_d$`Depth (m)`[], nb_d.tap, xlab='Depth(meters)',
#      ylab='', type='l', col='red', lwd=1,
#      main='Demeaned + Filtered Input Data', ylim=c(-8,8)
# )

```

# Step 5. Fourier Spectral Analysis
```{r}
#________________________________________________________________________________#
plot.frequency.spectrum <- function(X.k, 
                                    xlimits=c(0,length(X.k)/2), 
                                    xlab='Frequency',
                                    ylab='Amplitude',
                                    density=FALSE,
                                    plot.type='l',
                                    plot.period=FALSE
                                    ) { 
    x <- (0:(length(X.k)-1)) 
    plot.data  <- cbind(x, Mod(X.k))
  # TODO: why this scaling is necessary?
    plot.data[2:length(X.k),2] <- 2*plot.data[2:length(X.k),2] 
    
    N <- length(x)
    freq <- x
    
    amp <- plot.data[,2]
    if(density) {
      tot <- sum(amp)
      amp <- amp/tot
      print(amp)
      ylim <- c(0, max(amp))
    }
    else {
      tot <- 1
      ylim=c(0,max(Mod(plot.data[,2])))
    }
    
    plot(freq, amp, lwd=2, main="", type=plot.type,
                #xaxp=c(0, N, N/6),
                xlim=xlimits, # show only half of the frequency 
                ylim=ylim,
                xlab=xlab, ylab="Amplitude") 
    
    if (plot.period) {
      period <- (max_dr - min_dr) /freq
      plot(period, amp, lwd=2, main="", type=plot.type,
                #xaxp=c(0, N, N/6),
                #xlim=xlimits, # show only half of the frequency 
                #ylim=ylim,
                xlab="Period", 
                ylab="Amplitude") 
    }
    
}

nb_d.fft <- fft(nb_d.tap)
#par(mfrow=c(2,1))

xlab = sprintf('Frequency (cycle per %f meters)', max_d - min_d)
plot.frequency.spectrum(nb_d.fft,
                        xlim=c(0,200),
                        xlab=xlab,
                        ylab='Amplitude',
                        plot.period=TRUE#,
                        #,plot.type = 'h'
)
```

# With confidence interval
```{r}
plot.frequency.spectrum(nb_d.fft,
                        xlim=c(0,200),
                        xlab=xlab,
                        ylab='Amplitude'#,
                        #,plot.type = 'h'
)
n <- length(nb_d.fft)
R.f <- Mod(nb_d.fft)
I.f <- n * (R.f^2) 
var.I.f <- var(I.f)

N <- length(R.f)
freq <- 1:N
df <- 1/sqrt(12)
lines(freq, R.f/qchisq(0.975, df), t='l', lty=2, col='green')
par(new=T)
plot(freq, R.f/qchisq(0.025, df), t='l', lty=2, col='red', xlim=c(0, 200)
     , xaxt=NULL, yaxt=NULL, axes=FALSE, xlab="", ylab="")
```

```{r Step Power Spectrum + Calculate Wavelength}
# 
# Calculate Amplitude
Amp <- R.f <- Mod(nb_d.fft)

# Calculate Power Spectrum
PowerSpec <- Amp^2

# Take half of it , coz mirrored
PowerSpec <- PowerSpec[1:(length(Amp)/2)]

PowerSpec.han <- c()
i <- 1
nr <- length(PowerSpec)
for (n in 2:(nr-1)) {
  h <- hanning_ma(PowerSpec, n)
  PowerSpec.han <- c(PowerSpec.han, h)
}
dp <- as.numeric(nb_d$`Depth (m)`)
window_width <- max(dp) - min(dp)
wavenum <- 1:(length(PowerSpec.han))
wavelen <- window_width/wavenum

TotPowerSpec <- sum(PowerSpec)
RelPowerSpec <- PowerSpec.han/TotPowerSpec

# Charts of Wavenumber vs. Smoothed Power 
plot(wavenum, PowerSpec.han, t='h', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Power Spectra [Amplitude]^2'
)
plot(wavenum, PowerSpec.han, t='l', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Power Spectra [Amplitude]^2'
)

plot(wavenum, PowerSpec.han, t='h', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Power Spectra [Amplitude]^2',
     xlim=c(0,50))
plot(wavenum, PowerSpec.han, t='l', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Power Spectra [Amplitude]^2',
     xlim=c(0,50))

# Charts of Wavenumber vs. Relative Power 
plot(wavenum, RelPowerSpec, t='h', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Relative Power Spectra [Amplitude]^2')
# Charts of Wavenumber vs. Relative Power 
plot(wavenum, RelPowerSpec, t='l', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Relative Power Spectra [Amplitude]^2')
plot(wavenum, RelPowerSpec, t='h', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Relative Power Spectra [Amplitude]^2',
     xlim=c(0,120))
# Charts of Wavenumber vs. Relative Power 
plot(wavenum, RelPowerSpec, t='l', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Relative Power Spectra [Amplitude]^2',
     xlim=c(0,120))
# Charts of Wavenumber vs. Relative Power 
plot(wavenum, RelPowerSpec, t='h', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Relative Power Spectra [Amplitude]^2',
     xlim=c(0,50))
# Charts of Wavenumber vs. Relative Power 
plot(wavenum, RelPowerSpec, t='l', 
     xlab=paste('Wavenumber (cycle/window) window=', window_width, "m"), 
     ylab='Relative Power Spectra [Amplitude]^2',
     xlim=c(0,50))
```

# Wavelength
```{r Wavelength plotting}
# )Wavelength vs. Smoothed Power Spectra
plot(wavelen, PowerSpec.han, t='h', 
     xlab='Wavelength (m)', 
     ylab='Power Spectra [Amplitude]^2')

plot(wavelen, PowerSpec.han, t='l', 
     xlab='Wavelength (m)', 
     ylab='Power Spectra [Amplitude]^2')

plot(wavelen, PowerSpec.han, t='h', 
     xlab='Wavelength (m)', 
     ylab='Power Spectra [Amplitude]^2',
     xlim = c(0,40))
plot(wavelen, PowerSpec.han, t='h', 
     xlab='Wavelength (m)', 
     ylab='Power Spectra [Amplitude]^2',
     xlim = c(0,10), ylim=c(0,2000))

plot(wavelen, PowerSpec.han, t='l', 
     xlab='Wavelength (m)', 
     ylab='Power Spectra [Amplitude]^2',
     xlim = c(0,10), ylim=c(0,2000))
plot(wavelen, PowerSpec.han, t='l', 
     xlab='Wavelength (m)', 
     ylab='Power Spectra [Amplitude]^2',
     xlim = c(0,100), ylim=c(0,10000))

# )Wavelength vs. Relative Power Spectra
plot(wavelen, RelPowerSpec, t='h', 
     xlab='Wavelength (m)', 
     ylab='Relative Power Spectra [Amplitude]^2',
     xlim=c(0,10))

abline(v=15, lty=2)
d <- data.frame(WaveLength=wavelen, RelativePowerSpectra=RelPowerSpec)
ord <- order(d['RelativePowerSpectra'], decreasing=TRUE)
RelPowerSpec.sorted <- d[ord,]
RelPowerSpec.sorted$`WaveNumber(WindowWidth=601.68m)` <- wavenum[ord]

summary(nb_d)
head(nb_d)
tail(nb_d)
head(RelPowerSpec.sorted, 30)
```

```{r Step Calculation of Sedimentation Rate}
#________________________________________________________________________________#
### 
# Method #1:  Assuming a peak is a Milankovitch Cycle;  computing Sedimentation Rate
#Eccentricity   412, 123,  and 95 k.y.
#Obliquity      41 k.y.
#Precession    23, and 19 k.y.

#Long-period Peak a = 300.84m (2)
#Medium-period Peak b = 75.21 m (8)
#Medium-period Peak c = 23.14m (26)
#Short-period Peak d = 17.19m (35)
#Average of adjacent short -period Peaks e & f = 5.4 m (110, 111)
#Short -period Peak g = 4.6m (130, 131)

hm <- c(8, 35, 26, 2, 110, 131)
mil_cycles <- c(412, 123, 95, 41, 23, 19)

eccentricity <- 123

sed_rate <- (RelPowerSpec.sorted$WaveLength[1]/3) / (mil_cycles/1000)
sed_rate
# Unit: m/Ma

# a more accurate sedimentation rate can be computed.
```

```{r}
plot.harmonic <- function(X.k, i, ts, acq.freq, color="red", xlab='time', add=TRUE) {
    X.k.h <- rep(0,length(X.k))
    X.k.h[i+1] <- X.k[i+1] # i-th harmonic
    harmonic.trajectory <- get.trajectory(X.k.h, ts, acq.freq=acq.freq)
    if (add == TRUE)
      points(ts, harmonic.trajectory, type="l", col=color)
    else
      plot(ts, harmonic.trajectory, type="l", col=color, xlab=xlab)
    
    return(harmonic.trajectory) 
}
# returns the x.n time series for a given time sequence (ts) and
# a vector with the amount of frequencies k in the signal (X.k)
get.trajectory <- function(X.k,ts,acq.freq) {
  N   <- length(ts)
  i   <- complex(real = 0, imaginary = 1)
  x.n <- rep(0,N)           # create vector to keep the trajectory
  f  <- 0:(length(X.k)-1)
  
  # Equation  x_n = \frac{1}{N} sum_{0}^{N-1} (X_k * exp(i * 2 * pi * \frac{k * n}{ N })) 
  # X.k : amount of frequency k in the signal, each k-th value is a complex number including strength(amplitude) and phase shift
  # N   : number of samples
  # n   : current sample
  # k   : current frequency, between 0 Hz to N-1 Hz
  # 1/N : not necessary but it gives the actual sizes of the time spikes
  # n/N : percent of time we have gone through
  # 2pik: the speed in radians/second

  for(n in 0:(N-1)) {       # compute each time point x_n based on freqs X.k
    t <- n / N
    x.n[n+1] <- sum(X.k * exp(i*2*pi*f*t)) / N
  }
  
  x.n * acq.freq 
}

plot.show <- function(trajectory,
                      start_time = 0,
                      time=1, 
                      harmonics=-1, 
                      plot.freq=FALSE, 
                      plot.type='l',
                      scale=1.5,
                      xlab='Time',
                      ylab='') {
  acq.freq <- length(trajectory)/time      # data acquisition frequency (Hz)
  ts  <- seq(start_time,(start_time + time)-1/acq.freq,1/acq.freq) # vector of sampling time-points (s) 
  
  X.k <- fft(trajectory)
  x.n <- get.trajectory(X.k,ts, acq.freq=acq.freq) / acq.freq
  
  if (plot.freq)
    plot.frequency.spectrum(X.k, plot.type=plot.type)
  
  max.y <- ceiling(scale*max(Mod(x.n)))
  
  if (harmonics[1]==-1) {
    min.y <- floor(min(Mod(x.n)))-1
  } else {
    min.y <- ceiling(-scale*max(Mod(x.n)))
  }
  
  plot(ts,x.n, type="l", lwd=3, ylim=c(min.y,max.y), 
       xlab=xlab, ylab=ylab)
  #abline(h=min.y:max.y,v=0:time,lty=3, lwd=0.25)
  #points(ts,trajectory,cex=0.5, col="red")  # the data points we know
  
  y <- rep(0, length(X.k))
  if (harmonics[1]>-1) {
    for(i in 0:length(harmonics)) {
      r <- plot.harmonic(X.k, harmonics[i], ts, acq.freq, color=i+1)
      y <- y + r 
    }
  }
  points(ts, y/scale, t='l', col='orange', lwd=2)
  
}


amp <- Mod(nb_d.fft)
N <- length(amp)
amp.h <- amp[0:(N/2)]
x <- 0:(N/2-1)
d <- data.frame(x, amp.h)
d.sorted <- d[order(amp.h, decreasing = TRUE),]
head(d.sorted, 20)
```

# Let's take the 6 major cycles and try to reconstruct the original signal.
```{r}
hm <- c(8, 35, 26, 2, 110, 131)
plot.show(nb_d.tap, start_time = min_d, time = max_d - min_d, harmonics = hm,
          xlab='Depth (meter)', ylab='', scale=0.5)
hm_l <- paste(hm, ' cycles')
hm_l <- c(hm_l, 'combined cycle')
legend('topleft', legend=hm_l, col = c(1, 2, 3, 4, 5, 6,'darkblue'), 
       lty = rep(1,7), cex=0.40)

^```

# Output from R multi-taper library
```{r}
library(multitaper)
Mspec <- spec.mtm(nb_d.tap, deltat = 0.85, jackknife = TRUE,
                  k=7, nw=4,
                  log='no', lwd=2, 
                  xlim=c(0,0.25),
                  plot=FALSE)
x <- Mspec$freq
y <- Mspec$spec
plot(x, y, type='l', lwd=2
     ,xlim=c(0,0.2)
     #,log='y'
)
```

The frequencies are plotted in a scale between 0 to 0.5 (In our newark data 0 ~ 0 and 0.5 ~ ). The frequency resolution is low now. But we can still see the peaks around the same frequency.

#With Confidence interval.

```{r}
Mspec <- spec.mtm(nb_d.tap, deltat = 0.85, jackknife = TRUE,
                  k=7, nw=4,
                  log='no', lwd=2, 
                  xlim=c(0,0.25)
                  )

```

```{r}

plot(x, y, type='l', lwd=2
     ,xlim=c(0,0.2)
     ,log='y', main='Amplitudes ploted on logarithmic scale',
     xlab=seq(0,200)
)
```

# Output from R spectrum library
```{r}
Xspec <- spec.pgram(nb_d.tap, log='dB')
ylimits <- c(1/10*min(Xspec$spec), 25*max(Xspec$spec))
plot(Xspec$freq, Xspec$spec, xlim=c(0,0.2), type='l', lwd=2
     #, ylim=ylimits
)

df <- Xspec$df
lines(Xspec$freq, Xspec$spec/qchisq(0.975, df), t='l', lty=2, col='green')
lines(Xspec$freq, Xspec$spec/qchisq(0.025, df), t='l', lty=2, col='red')

bw <- Xspec$bandwidth

df <- Xspec$df
chi_lo <- df / qchisq(0.025, df)
chi_up <- df / qchisq(0.975, df)

#lines(Xspec$freq, Xspec$spec / chi_lo, col='green')
#par(new=T)
#plot(Xspec$freq, Xspec$spec * chi_up, xlim=c(0,0.2), col='red', type='l')

#plot(Xspec, xlim=c(0,0.2))
#plot(Xspec$freq, Xspec$spec, xlim=c(0,0.2), type='l', lwd=2)
#a <- qchisq(Xspec$spec * chi_up, df =Xspec$df)
#lines(Xspec$freq, pchisq(Xspec$spec, df=Xspec$df), col='red', xlim=c(0,0.2)) 
```

# R psd library output
```{r}
library(psd)
d <- ts(nb_d.tap, frequency=1)
Pspec <- psdcore(d, ntapaer=2)
plot(Pspec$freq, Pspec$spec, type='l', xlim=c(0,0.2))

spp <- spectral_properties(Pspec[["taper"]], db.ci = TRUE) 
psppu <- with(Pspec, create_poly(freq, dB(spec), spp$stderr.chi.upper))
#par(new=T)
plot(psppu$x.x, psppu$y.y, lty=2, type='l', 
     xlim=c(0, 0.2), col=c('green','red'),
     axes=FALSE, xlab='', ylab='')
```

```{r investigate lomb pacakge spectral methods}
library(lomb)
data(ibex)
lsp(ibex[2:3],)
lsp(ibex$temp,times=ibex$hours,type='period',ofac=5)
# lynx contains evenly sampled data
lsp(lynx)
lynx.spec <- lsp(lynx,type='period',from=2,to=20,ofac=5)
summary(lynx.spec)

lsp(as.numeric(nb_d$`Depth Rank`))
lsp(nb_d.dm, main='Lomb-Scargle Periodogram with demeaned data')
lsp(nb_d.tap, main='Lomb-Scargle Periodogram with demeaned + tapered data')
lsp(nb_d.tap, type='period')

data(ibex)
ibex.spec <- lsp(ibex[,2:3],type='period',from=12,to=36,ofac=10, plot=FALSE)
op <- par(pch=16)
plot.lsp(ibex.spec, main="Periodogram of daily rhythms of Tb in Capra ibex",
cex.lab=1.3,log="", type="b",level=FALSE,xaxt="n")
axis(side=1,at=seq(12,36,by=6))
par(op)

data(lynx)
set.seed(444)
rand.times <- sample(1:length(lynx),30) # select a random vector of sampling times
randlsp(1000,lynx[rand.times],times=rand.times)

randlsp(x =  nb_d.tap, type='period')
```

```{r investigate 'spectral' pacakge spectral methods}
library(spectral)
# create two sin-functions
x_orig <- seq(0,1,by=1e-2)
y_orig <- sin(10*2*pi*x_orig) + 0.3*sin(2*2*pi*x_orig)
# make a 10% gap
i <- round(length(x_orig)*0.2) : round(length(x_orig)*0.3)
x <- x_orig
y <- y_orig
x[i] <- NA
y[i] <- NA
# calculating the lomb periodogram

x <- as.numeric(nb_d$`Depth (m)`)
y <- as.numeric(nb_d.tap)
n <- length(x)
x <- x[2:(n-1)]
l <- spec.lomb(x = x, y = y)
# select a frequency range
m <- rbind(c(9,11))
# select and reconstruct the most significant component
l2 = filter.lomb(l, x, filt=m)
# plot everything
par(mfrow=c(2,1),mar = c(4,4,2,4))
plot(x,y,"l", main = "Gapped signal")
lines(l2$x, l2$y,lty=2)
legend("bottomleft",c("gapped","10Hz component"),lty=c(1,2))
plot(l,main = "Spectrum")
```

```{r investigate 'RobPer' pacakge spectral methods}
library(RobPer)
set.seed(7)
n <- 120
# equidistant time series:
zr <- tsgen(ttype="equi", ytype="const", pf=1, redpart= 0,
            s.outlier.fraction=0.2, 
            interval=FALSE, 
            npoints=n, 
            ncycles=n, 
            ps=1, 
            SNR=1, 
            alpha=1.5)

# Lomb-Scargle periodogram - Note: Due to the regression ansatz,
# RobPer is not able to compute period 2 in this case.
x <- as.numeric(nb_d$`Depth (m)`)
n <- length(x)
x <- x[2:(n-1)]
y <- as.numeric(nb_d.tap)
dat <- as.matrix(data.frame(x=x, y=y))
# periodogram of Fourier analysis
# y <- zr[,2]
PP_konv <- spec.pgram(y,#zr[,2], 
                      taper = 0,pad = 0, fast = FALSE, 
                      demean = TRUE,
                      detrend = TRUE, plot = TRUE)

PP_new <- RobPer(dat, #ts=zr, 
                 weighting=FALSE, 
                 periods=1/PP_konv$freq,
                 regression="tau", 
                 model="sine")
plot(PP_konv$freq, PP_konv$spec, 
     ylab="periodogram", xlab="frequency",
     main="Comparison of RobPer(...regression='LS', model='sine') and spec.pgram",
     cex=0.25)
points(PP_konv$freq, PP_new*var(y)*n/2, type="l")

legend("top",lty=c(1,0), pch=c(-5,1), legend=c("RobPer*var(y)*n/2", "spec.pgram"))
# Due to different ways of computation, the scaled periodograms are not exactly
# identical, but show very similar behavior.
```

```{r }


```
